Modelling BCG vaccination in the UK

### 2019-08-07 rmds/03-getTBinR.Rmd
thesis edits complete

`getTBinR` is an R package[@RCoreTeam2019] that I developed to facilitate working with the data[@WHO:2018] collected by the World Health Organization (WHO) on the country level epidemiology of Tuberculosis (TB). All data is freely available from the WHO^[WHO TB data: https://www.who.int/tb/country/data/download/en/] and the package code is archived on Zenodo[@Abbott:2019] and Github^[GitHub: https://github.com/seabbs/getTBinR/]. The aim of `getTBinR` is to allow researchers, and other interested individuals, to quickly and easily gain access to a detailed TB dataset and to start using it to derive key insights. It provides a consistent set of tools that can be used to rapidly evaluate hypotheses on a widely used dataset before they are explored further using more complex methods or more detailed data. Prior to the development of `getTBinR` access to the WHO data was ad-hoc and there were no standardised visualisation or summary tools.

>>>>

`getTBinR` is an R package[@RCoreTeam2019] that I developed to facilitate working with the data[@WHO:2018] collected by the World Health Organization (WHO) on the country level epidemiology of Tuberculosis (TB). All data is freely available from the WHO^[WHO TB data: https://www.who.int/tb/country/data/download/en/] and the package code is archived on Zenodo^[Zenodo: https://zenodo.org/badge/latestdoi/112591837] and Github^[GitHub: https://github.com/seabbs/getTBinR/]. The aim of `getTBinR` is to allow researchers, and other interested individuals, to quickly and easily gain access to a detailed TB dataset and to start using it to derive key insights. It provides a consistent set of tools that can be used to rapidly evaluate hypotheses on a widely used dataset before they are explored further using more complex methods or more detailed data. Prior to the development of `getTBinR` access to the WHO data was ad-hoc and there were no standardised visualisation or summary tools.

### 2019-08-07 rmds/09-model-fitting.Rmd
further formatting of figure positions
thesis edits complete

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. In order to use this model to understand TB transmission, and the impact of different BCG vaccination policies (Chapter \@ref(model-results)) this uncertainty needs to be reduced and the parameter space tightened to reflect more realistic ranges. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets. An alternative to model fitting is using a model parameterised with expert knowledge only for inference. This approach is not appropriate here due to the large amount of uncertainty for many of the model parameters. Any inference based in just the priors would have extremely large credible intervals, reflect reality poorly, and likely be biased in multiple areas.

@@

State space models (SSMs) may be used to model dynamic systems that are partially observed (such as all but the most contained infectious disease outbreak or endemic). They consist of a set of parameters, a latent continuous/discrete time state process and an observed continuous/discrete time process.[@Murray2015] The model developed in the previous section represents the state process of the SSM, with the parameters estimated for the model representing the model initial conditions and parameter set. To complete the mapping to an SSM an observational model is required. This observational model takes the latent estimates from the dynamic model and forecasts the observed data. Ispecify such an observational model in the Section \@ref(observational-model).

@@

Additional datasets were considered during initial model fitting and during later model calibration. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. Reducing the impact of stochastic noise, which makes fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approach used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were considered as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. Downsides of using these data are that reporting standards may have changed over time so a single measurement model may not be appropriate and non-UK born cases are included in these data making fitting to this data dependent on the number of non-UK born cases pre 2000 which are themselves estimated during model fitting.

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that under reporting is more likely that over reporting. The prior for the reporting noise is based on the observed variation between years.

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000. A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, this may result in overfitting and so has not been implemented here.

@@

The model outlined in Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful transmission related information in the observed data may be lost due to overfitting from other variables.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates.[@Murray2015] Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

`LibBi` was used for all model fitting.[@Murray2015] `LibBi` is a software package for state-space modelling and Bayesian inference. It uses a domain specific language for model specification, which is then optimised and compiled to provide highly efficient model code. It focuses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across multiple CPUs or GPUs. The `rbi` and `rbi.helpers` packages were used to interface with `LibBi` from R.[@Funk:2019ud; @Funk:2019uw] `rbi.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked (using CPU voltage manipulation), with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. Whilst a compute cluster was theoretically available, in practise the hardware available was limited, installing `LibBi` was challenging, and run times were constrained by fair access. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

In the previous chapter multiple potential model structures were outlined, each of which could be valid based on theoretical considerations. The observed data can be used to identify which of these structures best reflects reality. This can be done using the deviance information criterion (DIC). The DIC is a hierarchical modeling generalization of the Akaike information criterion (AIC) and can be used to compare nested models.[@Gelman:nll_LBlw]

@@

Varying the proposal size, rejuvenation threshold, and number of rejuvenation steps showed some promise at improving the quality of the model fit but ultimately computational constraints limited how much progress could be made using this approach. It is possible that a much longer run time could result in an improved fit to data with no other model changes. Alternatively these results may driven by the model being to complex for an SMC-SMC approach to be viable in this framework.

@@

Whilst none of the scenarios fitted the data well scenarios that included variable transmission probability in young adults fitted the data much better than those that did not (Table \@ref(tab:09-scenarios-table)). When considered on its own, allowing non-UK born mixing to vary for young adults had only a small impact on the quality of the model fit in comparison to allowing the transmission probability to vary. However, the scenario that allowed both the transmission probability and non-UK born mixing to vary fit the data much better than any other scenario considered (Table \@ref(tab:09-scenarios-table)). In the following sections only the results from this scenario will be discussed.

@@

The fitted model had a low effective sample size manifesting in a spuriously tight posterior distribution (Figure \@ref(fig:09-fig-prior-params); Table \@ref(tab:09-prior-posterior)). This indicates that model fitting did not allow for a full exploration of the parameter space. Given that a relatively tight proposal distribution was used this may indicate that an independent proposal is insufficient. It is likely that the key factor behind the models consistent under-prediction is the selection of an extremely low value for the current effective contact rate. It is difficult to interpret these findings further due the low quality of the posterior distribution.

@@

Figure \@ref(fig:09-fig-sens) shows the partial rank correlation coefficients for each parameter that was fitted too. This indicates that variation in the historic effective contact rate leads to the greatest variation in TB incidence. Based on the model structure this makes sense as small changes in the historic contact rate will compounded over the run time of the model. Similarly the half-life of the historic contact rate also plays a dominant role. The non-UK born mixing modifier for young adults leads to the second largest change in TB incidence. This corresponds to the findings from the scenario analysis discussed above in which the introduction of this parameter resulted in a greatly improved model fit. Surprisingly, the transmission probability modifier has relatively little impact on TB incidence compared to the other parameters fitted to, disagreeing with the findings from the scenario analysis. The lack of diversity in the posterior distribution seen above means that these findings are indicative only.

@@

In this chapter I have formulated the disease transmission model developed in the previous chapter as a state-space model, developed a model fitting pipeline to fit this model to observed data, discussed the approaches taken to try and improve the quality of the model fit, and presented preliminary results from the model fitting pipeline. The model fit the observed data poorly using the approach layed out here. Whilst multiple ad-hoc approaches were used to try and improve the quality of the model fit these did little to improve it. The model consistently under-predicted overall TB incidence and also failed to reproduce the observed age distribution. There was little evidence of a good fit to trends over time in the observed data. Model comparison showed that models that included modifying parameters for the transmission probability, and non-UK born mixing in young adults fitted the observed data better than those that did not. The estimated posterior distributions for all parameters were spuriously tight in comparison to the prior distributions. This may be evidence of poor mixing meaning that any conclusions drawn about the parameter sets selected may be incorrect.  Parameters that contributed to TB transmission across the time-span of the model dominated those that only contributed to recent TB transmission in the parameter sensitivity analysis.

As discussed earlier in this chapter, SMC-SMC was used for all model fitting. This uses an external SMC step to estimate the posterior distribution as well as an internal SMC step to estimate the marginal likelihood. In general Bayesian model fitting approaches are beneficial as they allow prior information to be fully incorporated into model fitting and they produce posterior distribution estimates rather than point estimates.[@Murray2015; @Toni2009] There are two main families of approaches, SMC and MCMC. Using SMC to estimate the posterior distribution has numerous advantages over MCMC. The first of these is that MCMC approaches are sensitive to their initial conditions. If a model has multiple local best fits MCMC may only converge to a single minima rather than fully exploring the posterior distribution. Multiple MCMC chains may be used to try and account for this but as each chain must be independently run to convergence only a few concurrent chains are likely to be practical. SMC on the other hand is initialised with a large sample from the prior distribution, meaning that local minimas are more likely to be explored. Parameter particle weighting and re-sampling then balances the contribution to the posterior distribution of these local minimas based on their fit to the observed data. Additionally, MCMC approaches are by definition sequential,[@Murray2015] although if they make use of particle filters these can be run in parallel. Increasing the number of particles in a filter may lead to an increase in the chain mixing rate of the MCMC chain but as particles numbers are increased any returns will decrease. To account for this multiple chains are often used, but as outlined above the burn-in required for each chain limits the potential speed-up. In comparison, each SMC parameter particle can have it's marginal likelihood computed separately. Although the re-sampling step remains a bottleneck as it can only be completed once all marginal likelihoods have been computed. On the other hand SMC is less interactive than MCMC meaning that model fitting is harder to inspect when it is in progress. This is because SMC is not sequential, unlike an MCMC run for which each draw can be inspected as it is computed. Similarly, as SMC is not a sequential technique multiple runs cannot be combined. This means that model fitting must be done in a single run using a priori knowledge to judge the number of MCMC rejuvenation steps required, and the expected total run time. SMC will also have a variable run time based on the effective sample size as rejuvenation only happens when parameter particles have been depleted beyond a certain point. An additional benefit of SMC approaches are that they can theoretically be extended to model selection as well as parameter posterior estimation - effectively estimating posteriors for candidate model structures.[@Toni2009] Beyond SMC and MCMC there are multiple other model fitting algorithms that each have their own strengths and benefits - discussion of these is outside the scope of this thesis.

The particle filter has been shown to provide an unbiased estimate of the likelihood for arbitrary state-space models.[@Murray2015] As a full information technique the particle filter provides a more accurate estimate of the likelihood than other approximate techniques, with relatively little tuning or user interaction.[@Murray2015] The major downside of the particle filter, compared to other approaches, is the high compute requirements, with each particle requiring a full model simulation. For highly complex models, the particle filter approach may not be tractable or a reduced level of accuracy of the marginal likelihood estimate must be accepted. In addition, the bootstrap particle filter may become depleted (i.e variance between particles is reduce to such an extent that the effective sample size becomes small) when the model being fitted is a poor fit for the observed data.[@Murray2015] Whilst this can be resolved using additional particles, or rejuvenation steps, this may not be computationally tractable. There are several alternative particle filters that seek to mitigate these issues but many of them are highly complex and do not significantly reduce the required compute - see [@Murray2015] for a detailed discussion of some of these alternatives. An alternative approach to the particle filter is to use an approximate technique, such as approximate Bayesian computation (ABC). ABC can be used to estimate the likelihood by comparing observed and simulated data.[@Toni2009] This dramatically reduces the required compute as multiple model simulations per parameter set are no longer required. The comparison between the observed and simulated data is facilitated using a distance function with parameter sets being accepted if they are within some threshold distance. This threshold can be be tuned to produce a good estimate of the posterior distribution. Developing a distance function for all of the observed data can often be challenging so instead the distance is often calculated using a set of summary features.[@Toni2009] In principle, ABC can be used with a wide variety of algorithms (including a simple rejection approach, MCMC, and SMC) to estimate the posterior distribution but in practice it has been shown that ABC-SMC generally performs better than alternative ABC approaches.[@Toni2009] The two major limitations of ABC compared to the use of a particle filter is that in most cases summary statistics must be used rather than calculating the distance from the observed data and a function for calculating the distance must be chosen.[@Toni2009; @Lintusaari2016] Whilst some techniques exist for evaluating summary statistics the choice is often subjective, relying on domain knowledge.[@Lintusaari2016] Chosen summary statistics rarely capture the information contained in the observed data fully and can inflate posterior distribution and in the worst cases introduce bias to the estimates.[@Busetto2013] The choice of distance function may also influence the estimates of the posterior distribution.[@Lintusaari2016]

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Indeed the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's `R` interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

@@

Whilst the results presented here are not encouraging it may be the case that with a greatly extended run-time a better fitting model may be found. The first step to testing this is to rerun the pipeline using several orders of magnitude more rejuvenation steps - as these have no RAM overhead. Even if the resulting model fits are still poor this may indicate areas for improvement - in either the model fitting pipeline or the model itself. Reducing model complexity may also help via decreasing the computational burden and so allowing more particles to be used. Possible simplifications are to remove features, such as the treatment population, that do little to alter the dynamics but are present due to use case concerns or switch from a continous framework to a discrete one. An alternative option is to re-code the model developed in the last chapter into a more generic form (i.e C) and to attempt to fit the model using other techniques that are less compute intensive and more robust - such as SMC-ABC. Another option would be to re-implement the model in the form required by another modelling package - such as `pomp`. However, this would mean running the risk of again getting stuck within a framework that is difficult to debug and may not be working as expected. A final option would be to reduce the complexity of the model and potentially fit to less complex data. This may be the only solution if the current model is not identifiable.

>>>>

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. In order to use this model to understand TB transmission, and the impact of different BCG vaccination policies (Chapter \@ref(model-results)) this uncertainty needs to be reduced and the parameter space tightened to reflect more realistic ranges. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets. An alternative to model fitting is using a model parameterised with expert knowledge only for inference. This approach is not appropriate here due to the large amount of uncertainty for many of the model parameters. Any inference based on just the parametisation from the previous chapter would have large credible intervals, reflect reality poorly, and likely be biased in multiple areas.

@@

State space models (SSMs) may be used to model dynamic systems that are partially observed (such as all but the most contained infectious disease outbreak or endemic). They consist of a set of parameters, a latent continuous/discrete time state process and an observed continuous/discrete time process.[@Murray2015] The model developed in the previous section represents the state process of the SSM, with the parameters estimated for the model representing the model initial conditions and parameter set. To complete the mapping to an SSM an observational model is required. This observational model takes the latent estimates from the dynamic model and forecasts the observed data. I specify such an observational model in Section \@ref(observational-model).

@@

Additional datasets were considered during initial model fitting and during later model calibration. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. This reduced the impact of stochastic noise, making fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approach used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were considered as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. Downsides of using these data are that reporting standards may have changed over time so a single measurement model may not be appropriate and non-UK born cases are included in these data making fitting to this data dependent on the number of non-UK born cases pre 2000 which are themselves estimated during model fitting.

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that underreporting is more likely than over-reporting. The prior for the reporting noise is based on the observed variation between years.

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000. A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, this may result in over-fitting and so has not been implemented here.

@@

The model outlined in Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful transmission related information in the observed data may be lost due to over-fitting from other variables.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates.[@Murray2015] Unfortunately, many of these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

`LibBi` was used for all model fitting.[@Murray2015] `LibBi` is a software package for state-space modelling and Bayesian inference. It uses a domain specific language for model specification, which is then optimised and compiled to provide highly efficient model code. It focuses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across multiple CPUs or GPUs. The `rbi` and `rbi.helpers` packages were used to interface with `LibBi` from R.[@Funk:2019ud; @Funk:2019uw] `rbi.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked (using CPU voltage manipulation), with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. Whilst a cluster was theoretically available, in practise the hardware available was limited, installing `LibBi` was challenging, and run times were constrained by fair access. All model fitting code is available on GitHub as an R package^[Model fitting pipeline R package: https://github.com/seabbs/ModelTBBCGEngland].

@@

In the previous chapter multiple scenarios were outlined, each of which could be valid based on theoretical considerations. The observed data can be used to identify which of these scenarios best reflects reality. This can be done using the deviance information criterion (DIC). The DIC is a hierarchical modeling generalization of the Akaike information criterion (AIC) and can be used to compare nested models.[@Gelman:nll_LBlw]

@@

Varying the proposal size, rejuvenation threshold, and number of rejuvenation steps showed some promise at improving the quality of the model fit but ultimately computational constraints limited how much progress could be made using this approach. It is possible that a much longer run time could result in an improved fit to data with no other model changes. Alternatively these results may be driven by the model being to complex for an SMC-SMC approach to be viable.

@@

Whilst none of the scenarios fitted the data well, scenarios that included variable transmission probability in young adults fitted the data much better than those that did not (Table \@ref(tab:09-scenarios-table)). When considered on its own, allowing non-UK born mixing to vary for young adults had only a small impact on the quality of the model fit in comparison to allowing the transmission probability to vary. However, the scenario that allowed both the transmission probability and non-UK born mixing to vary fit the data much better than any other scenario considered (Table \@ref(tab:09-scenarios-table)). In the following sections only the results from this scenario will be discussed.

@@

The fitted model had a low effective sample size manifesting in  spuriously tight posterior distributions (Figure \@ref(fig:09-fig-prior-params); Table \@ref(tab:09-prior-posterior)). This indicates that model fitting did not allow for a full exploration of the parameter space. Given that a relatively tight proposal distribution was used this may indicate that an independent proposal is insufficient. It is likely that the key factor behind the models consistent under-prediction is the selection of a low value for the current effective contact rate. It is difficult to interpret these findings further due the low quality of the posterior distribution.

@@

Figure \@ref(fig:09-fig-sens) shows the partial rank correlation coefficients for each parameter that was fitted to. It indicates that variation in the effective contact rate and non-UK born mixing lead to the greatest variation in TB incidence. Based on the model structure this makes sense as these paremeters are directly linked to modern day transmission. The parameters that modify TB transmission and non-UK born mixing in young adults also lead to significant variation in TB incidence. This corresponds to the findings from the scenario analysis discussed above in which the introduction of these parameters resulted in a greatly improved model fit. The lack of diversity in the posterior distribution seen above means that these findings are indicative only.

@@

In this chapter I have formulated the disease transmission model developed in the previous chapter as a state-space model, developed a model fitting pipeline to fit this model to observed data, discussed the approaches taken to try and improve the quality of the model fit, and presented preliminary results from the model fitting pipeline. The model fit the observed data poorly using the approach layed out here. Whilst multiple ad-hoc approaches were used to try and improve the quality of the model fit these did little to improve it. The model consistently under-predicted overall TB incidence and also failed to reproduce the observed age distribution. There was little evidence of a good fit to trends over time in the observed data. Model comparison showed that models that included modifying parameters for the transmission probability, and non-UK born mixing in young adults fitted the observed data better than those that did not. The estimated posterior distributions for all parameters were spuriously tight in comparison to the prior distributions. This may be evidence of poor mixing meaning that any conclusions drawn about the parameter sets selected may be incorrect. Parameters that contributed to recent TB transmission dominated those that contributed across the time-span of the model in the parameter sensitivity analysis.

As discussed earlier in this chapter, SMC-SMC was used for all model fitting. This uses an external SMC step to estimate the posterior distribution as well as an internal SMC step to estimate the marginal likelihood. In general, Bayesian model fitting approaches are beneficial as they allow prior information to be fully incorporated into model fitting and they produce posterior distribution estimates rather than point estimates.[@Murray2015; @Toni2009] There are two main families of approaches, SMC and MCMC. Using SMC to estimate the posterior distribution has numerous advantages over MCMC. The first of these is that MCMC approaches are sensitive to their initial conditions. If a model has multiple local best fits MCMC may only converge to a single minima rather than fully exploring the posterior distribution. Multiple MCMC chains may be used to try and account for this but as each chain must be independently run to convergence only a few concurrent chains are likely to be practical. SMC on the other hand is initialised with a large sample from the prior distribution, meaning that local minimas are more likely to be explored. Parameter particle weighting and re-sampling then balances the contribution to the posterior distribution of these local minimas based on their fit to the observed data. Additionally, MCMC approaches are by definition sequential,[@Murray2015] although if they make use of particle filters these can be run in parallel. Increasing the number of particles in a filter may lead to an increase in the chain mixing rate of the MCMC chain but as particles numbers are increased any returns will decrease. To account for this multiple chains are often used, but as outlined above the burn-in required for each chain limits the potential speed-up. In comparison, each SMC parameter particle can have it's marginal likelihood computed separately. Although the re-sampling step remains a bottleneck as it can only be completed once all marginal likelihoods have been computed. On the other hand SMC is less interactive than MCMC meaning that model fitting is harder to inspect when it is in progress. This is because SMC is not sequential, unlike an MCMC run for which each draw can be inspected as it is computed. Similarly, as SMC is not a sequential technique multiple runs cannot be combined. This means that model fitting must be done in a single run using a priori knowledge to judge the number of MCMC rejuvenation steps required, and the expected total run time. SMC will also have a variable run time based on the effective sample size as rejuvenation only happens when parameter particles have been depleted beyond a certain point. An additional benefit of SMC approaches are that they can theoretically be extended to model selection as well as parameter posterior estimation - effectively estimating posteriors for candidate model structures.[@Toni2009] Beyond SMC and MCMC there are multiple other model fitting algorithms that each have their own strengths and benefits - discussion of these is outside the scope of this thesis.

The particle filter has been shown to provide an unbiased estimate of the likelihood for arbitrary state-space models.[@Murray2015] As a full information technique the particle filter provides a more accurate estimate of the likelihood than other approximate techniques with relatively little tuning or user interaction.[@Murray2015] The major downside of the particle filter, compared to other approaches, is the high compute requirements, with each particle requiring a full model simulation. For highly complex models, the particle filter approach may not be tractable or a reduced level of accuracy of the marginal likelihood estimate must be accepted. In addition, the bootstrap particle filter may become depleted (i.e variance between particles is reduce to such an extent that the effective sample size becomes small) when the model being fitted is a poor fit for the observed data.[@Murray2015] Whilst this can be resolved using additional particles, or rejuvenation steps, this may not be computationally tractable. There are several alternative particle filters that seek to mitigate these issues but many of them are highly complex and do not significantly reduce the required compute - see [@Murray2015] for a detailed discussion of some of these alternatives. An alternative approach to the particle filter is to use an approximate technique, such as approximate Bayesian computation (ABC). ABC can be used to estimate the likelihood by comparing observed and simulated data.[@Toni2009] This dramatically reduces the required compute as multiple model simulations per parameter set are no longer required. The comparison between the observed and simulated data is facilitated using a distance function with parameter sets being accepted if they are within some threshold distance. This threshold can be be tuned to produce a good estimate of the posterior distribution. Developing a distance function for all of the observed data can often be challenging so instead the distance is often calculated using a set of summary features.[@Toni2009] In principle, ABC can be used with a wide variety of algorithms (including a simple rejection approach, MCMC, and SMC) to estimate the posterior distribution but in practice it has been shown that ABC-SMC generally performs better than alternative ABC approaches.[@Toni2009] The two major limitations of ABC compared to the use of a particle filter is that in most cases summary statistics must be used rather than calculating the distance from the observed data and a function for calculating the distance must be chosen.[@Toni2009; @Lintusaari2016] Whilst some techniques exist for evaluating summary statistics the choice is often subjective, relying on domain knowledge.[@Lintusaari2016] Chosen summary statistics rarely capture the information contained in the observed data fully and can inflate posterior distribution and in the worst cases introduce bias to the estimates.[@Busetto2013] The choice of distance function may also influence the estimates of the posterior distribution.[@Lintusaari2016]

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Alternatively, the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's `R` interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

@@

Whilst the results presented here are not encouraging it may be the case that with a greatly extended run-time a better fitting model may be found. The first step to testing this is to rerun the pipeline using several orders of magnitude more rejuvenation steps - as these have no RAM overhead. Even if the resulting model fits are still poor this may indicate areas for improvement - in either the model fitting pipeline or the model itself. Reducing model complexity may also help via decreasing the computational burden, allowing more particles to be used, and also reducing the size of the potential parameter space. Possible simplifications are to remove features, such as the treatment population, that do little to alter the dynamics but are present due to use case concerns or switch from a continous framework to a discrete one. An alternative option is to re-code the model developed in the last chapter into a more generic form (i.e C) and to attempt to fit the model using other techniques that are less compute intensive and more robust - such as SMC-ABC. Another option would be to re-implement the model in the form required by another modelling package - such as `pomp`. However, this would mean running the risk of again getting stuck within a framework that is difficult to debug and may not be working as expected. A final option would be to reduce the complexity of the model and potentially fit to less complex data. This may be the only solution if the current model is not identifiable.
